\PYGEOMETRY{}: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2021-03-28 public/paper/sub-cpc/pyg4ometry.tex
Update to archive and CPC author statement

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\linenumbers

@@

\begin{figure*}[hbt!]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Schematic of \PYGEOMETRY{} workflow, showing (a) the different input file formats, (b) Python processing, (c) output file targets and
  (d) MCRT codes which use the geometry. }
\end{figure*}

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[width=5cm]{./diagrams/lvToPv.pdf}
\caption{The placement of a physical volume inside a logical volume.}
\label{fig:lvToPv}
\end{center}
\end{figure}

@@

\subsection{Overlap detection}
All MCRT codes cannot handle spatial overlap between two geometric objects and will have ill-defined behaviour when tracking particles
in such a situation.  A key feature of \PYGEOMETRY{} is the detection of potential overlaps in a way which is most useful to the user, it does this by
performing an intersection operation between solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows three different types of possible
overlaps, (a) protrusion of a daughter from the mother, (b)  finite volume intersection between two daughters  and (c) an overlap where two daughters
share a face. If the resulting intersection is non-null then the overlaps can be displayed side-by-side in the visualisation.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between the daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\columnwidth]{./diagrams/rapidModelling.pdf}
\caption{VTK visualisation output from code Listing~\ref{lst:pythonRapidModelling}.}
\label{fig:rapidModellingExample}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\columnwidth]{./model-scene/faradayCup2.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML. Above:
  the original FLUKA geometry displayed in flair, FLUKA's graphical user interface.
  Below: the GDML geometry viewed using \PYGEOMETRYPOS{} VTK visualiser. The example is a Faraday cup used to capture
and measure accelerator beam charge.}
\label{fig:fluka-to-geant4-cup}
\end{center}
\end{figure}

@@

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=6cm]{./diagrams/lattice}
\caption{The lattice feature demonstrating a lattice cell
  referring to its basic unit with the rototranslation~$R_\textrm{trans}$.
  Any particle entering the cell will be transformed onto the basic unit
  with $R_\textrm{trans}$, and when leaving the basic unit, back to the cell
  with $R_\textrm{trans}^{-1}$.}
\label{fig:lattice}
\end{center}
\end{figure}
This feature is clearly analogous to the logical/physical volume feature in
Geant4, although it is more implicit as the contents of a given logical
volume are explicitly stated, whereas the contents of a lattice cell
are implied by the combination of the rototranslation and the global positions of
its lattice cell and basic unit.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each
of the Geant4 solids can be mapped to a FLUKA region.  A region is a volume
of space defined by a material and the Boolean disjunction (a union using
the operator~$\:|\:$ in free format geometry, the most widely-used FLUKA
CSG format) of one or more zones. Each zone is then defined in terms
of the conjunction (intersection with~$+$, subtraction with~$-$) of one or
more primitive bodies, as well as parentheses to determine the order of
operations within the zone.  FLUKA has 20 of these primitive bodies, listed
in Table~\ref{tab:Fluka2Geant4} and, in general, infinite-extent bodies
have tracking accuracy and efficiency benefits over finite ones. Key for
conversion are XY-, XZ-, YZ-Planes (XYP, XZP, YZP), arbitrary plane (PLA),
Z-axis aligned cylinder (ZCC), Z-axis aligned elliptical cylinder (ZEC),
sphere (SPH), truncated right-angle cone (TRC) and general quadric surface
(QUA). Some solids in Geant4 directly map to a single FLUKA body, others
require the construction of a simple FLUKA CSG tree combining these
primitives. Table~\ref{tab:geant2fluka} lists the mapping between Geant4
solids and the bodies used to compose a FLUKA region.
\begin{figure}
\begin{center}
\includegraphics[width=0.9\columnwidth]{./model-scene/CuboidalChamber.pdf}
\caption{Example conversion of a simple GDML geometry to FLUKA, above:
the original model in \PYGEOMETRY{}, below: the converted FLUKA geometry
viewed in flair. The example is a sector bend dipole electromagnet.}
\label{fig:gdml-flair}
\end{center}
\end{figure}

@@

\subsection{Disjunctive normal form and degenerate surfaces}
FLUKA will typically decompose a region into disjunctive normal form (DNF)
at runtime, this normal form is characterised as the union of intersections and subtractions,
\begin{equation}
R = z_1 \; | \;z_2\;  | \; z_3 	\; | \; z_4 \dots
\end{equation}
Defining regions in terms of the DNF allows the rapid test of whether a point is inside that region. Testing each zone
$z_i$ of $R$ can terminate if a point is determined to be inside any of the zones $z_i$. In general
Equation~\ref{eqn:logicalSubtraction} does not have the form of a DNF. If there are many levels of logical-physical volume
placement, then recursive application of  Equation~\ref{eqn:logicalSubtraction} will create a nested set of
parentheses. There are some conditions where a general Boolean expression can yield an exponential
explosion of the final DNF. There are well known algorithms to convert logical expressions to its DNF.
\PYGEOMETRY{} can simplify parentheses from a region by creating a corresponding SymPy~\cite{10.7717/peerj-cs.103} Boolean
expression and using the \verb|to_dnf| method.  Further simplification of the CSG tree leverages \PYGEOMETRY{}'s
meshing capabilities combined with CSG pruning algorithms based
on~\cite{pruning}.  FLUKA will by default try to expand all regions to
their DNF at runtime, which inevitably can result in the sort of exponential explosion already mentioned.  Until version 4.0, if FLUKA identified
such an explosion, it would the expansion, report an error and exit, thus making such models impossible to run.  As of version 4.0, however, this
expansion can be disabled, and the tracking algorithms will walk the CSG trees verbatim, albeit at some tracking efficiency cost.  Therefore, most
output generated from the GDML to FLUKA conversion described here can only be used with FLUKA 4.0.

@@

\section{STL and CAD to GDML conversion}
STL and CAD conversion are closely related. In both cases the solid
(in the case of STL) and solids (in the case of CAD) are converted to
tessellated solid(s). STL is a relatively simple file format that can
be loaded using pure Python. As STL files typically only contain a single
solid the \pyinline{stl.Reader} provides a single solid and not a logical volume
as with the other file readers.
%
\begin{figure}
\begin{center}
\includegraphics[width=0.9\columnwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple CAD (STEP) geometry (sector bend
  dipole electromagnet) to GDML, above: the original model in FreeCAD,
  below: the GDML geometry viewed using \PYGEOMETRY{}. }
\label{fig:cad-gdml}
\end{center}
\end{figure}

@@

\section{Complete simulation example}
\PYGEOMETRY{} is designed to be as flexible as possible and offer the user a wide range of usage styles, input files
and workflows. A fictitious beamline was created to demonstrate the capabilities of \PYGEOMETRY, this creates a composite
\emph{scene} which consists of geometry sources from the different formats described in this paper. The beamline consists of a vacuum chamber
(modelled in \PYGEOMETRY{}), a vacuum gate valve (STL from the manufacturer), a triplet of quadrupole magnets (exported to GDML
from BDSIM), a sector bend dipole electromagnet (created in Autodesk Fusion 360) and finally a Faraday cup (FLUKA geometry designed in
flair). Each different file is loaded using the appropriate \PYGEOMETRY{} \pyinline{Reader} class and then placed as a physical volume. The final composite
geometry is shown in Figure~\ref{fig:model}. It must be noted that when this complete geometry is written to GDML and loaded into
Geant4 it cannot be visualised with anything but the ray tracer due to limitations in the OpenGL visualisation in Geant4. Another
important note is that this geometry cannot be converted to FLUKA as it contains tessellated solids (both the STL gate valve and
dipole magnet).
%
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/model.pdf}
\caption{Complete composite example using \PYGEOMETRY{} showing a model with geometry sourced from a range of different formats.}
\label{fig:model}
\end{center}
\end{figure*}

@@

>>>>

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}
%\documentclass[preprint,12pt]{elsarticle}

@@

%\linenumbers

@@

\begin{figure*}[hbt!]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{workflow.pdf}
  \caption{\label{fig:workflow}Schematic of \PYGEOMETRY{} workflow, showing (a) the different input file formats, (b) Python processing, (c) output file targets and
  (d) MCRT codes which use the geometry. }
\end{figure*}

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[width=5cm]{lvToPv.pdf}
\caption{The placement of a physical volume inside a logical volume.}
\label{fig:lvToPv}
\end{center}
\end{figure}

@@

\subsection{Overlap detection}
All MCRT codes cannot handle spatial overlap between two geometric objects and will have ill-defined behaviour when tracking particles
in such a situation.  A key feature of \PYGEOMETRY{} is the detection of potential overlaps in a way which is most useful to the user, it does this by
performing an intersection operation between solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows three different types of possible
overlaps, (a) protrusion of a daughter from the mother, (b)  finite volume intersection between two daughters  and (c) an overlap where two daughters
share a face. If the resulting intersection is non-null then the overlaps can be displayed side-by-side in the visualisation.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{overlap}
\caption{Schematic of the three different types of overlap between the daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\columnwidth]{rapidModelling}
\caption{VTK visualisation output from code Listing~\ref{lst:pythonRapidModelling}.}
\label{fig:rapidModellingExample}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\columnwidth]{faradayCup2.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML. Above:
  the original FLUKA geometry displayed in flair, FLUKA's graphical user interface.
  Below: the GDML geometry viewed using \PYGEOMETRYPOS{} VTK visualiser. The example is a Faraday cup used to capture
and measure accelerator beam charge.}
\label{fig:fluka-to-geant4-cup}
\end{center}
\end{figure}

@@

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=6cm]{lattice}
\caption{The lattice feature demonstrating a lattice cell
  referring to its basic unit with the rototranslation~$R_\textrm{trans}$.
  Any particle entering the cell will be transformed onto the basic unit
  with $R_\textrm{trans}$, and when leaving the basic unit, back to the cell
  with $R_\textrm{trans}^{-1}$.}
\label{fig:lattice}
\end{center}
\end{figure}
This feature is clearly analogous to the logical/physical volume feature in
Geant4, although it is more implicit as the contents of a given logical
volume are explicitly stated, whereas the contents of a lattice cell
are implied by the combination of the rototranslation and the global positions of
its lattice cell and basic unit.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each
of the Geant4 solids can be mapped to a FLUKA region.  A region is a volume
of space defined by a material and the Boolean disjunction (a union using
the operator~$\:|\:$ in free format geometry, the most widely-used FLUKA
CSG format) of one or more zones. Each zone is then defined in terms
of the conjunction (intersection with~$+$, subtraction with~$-$) of one or
more primitive bodies, as well as parentheses to determine the order of
operations within the zone.  FLUKA has 20 of these primitive bodies, listed
in Table~\ref{tab:Fluka2Geant4} and, in general, infinite-extent bodies
have tracking accuracy and efficiency benefits over finite ones. Key for
conversion are XY-, XZ-, YZ-Planes (XYP, XZP, YZP), arbitrary plane (PLA),
Z-axis aligned cylinder (ZCC), Z-axis aligned elliptical cylinder (ZEC),
sphere (SPH), truncated right-angle cone (TRC) and general quadric surface
(QUA). Some solids in Geant4 directly map to a single FLUKA body, others
require the construction of a simple FLUKA CSG tree combining these
primitives. Table~\ref{tab:geant2fluka} lists the mapping between Geant4
solids and the bodies used to compose a FLUKA region.
\begin{figure}
\begin{center}
\includegraphics[width=0.9\columnwidth]{CuboidalChamber}
\caption{Example conversion of a simple GDML geometry to FLUKA, above:
the original model in \PYGEOMETRY{}, below: the converted FLUKA geometry
viewed in flair. The example is a sector bend dipole electromagnet.}
\label{fig:gdml-flair}
\end{center}
\end{figure}

@@

\subsection{Disjunctive normal form and degenerate surfaces}
FLUKA will typically decompose a region into disjunctive normal form (DNF)
at runtime, this normal form is characterised as the union of intersections and subtractions,
\begin{equation}
R = z_1 \; | \;z_2\;  | \; z_3 	\; | \; z_4 \dots
\end{equation}
Defining regions in terms of the DNF allows the rapid test of whether a point is inside that region. Testing each zone
$z_i$ of $R$ can terminate if a point is determined to be inside any of the zones $z_i$. In general
Equation~\ref{eqn:logicalSubtraction} does not have the form of a DNF. If there are many levels of logical-physical volume
placement, then recursive application of  Equation~\ref{eqn:logicalSubtraction} will create a nested set of
parentheses. There are some conditions where a general Boolean expression can yield an exponential
explosion of the final DNF. There are well known algorithms to convert logical expressions to its DNF.
\PYGEOMETRY{} can simplify parentheses from a region by creating a corresponding SymPy~\cite{10.7717/peerj-cs.103} Boolean
expression and using the \verb|to_dnf| method.  Further simplification of the CSG tree leverages \PYGEOMETRY{}'s
meshing capabilities combined with CSG pruning algorithms based
on~\cite{pruning}.  FLUKA will by default try to expand all regions to
their DNF at runtime, which inevitably can result in the sort of exponential explosion already mentioned.  Until version 4.0, if FLUKA identified
such an explosion, it would terminate the expansion, report an error and exit, thus making such models impossible to run.  As of version 4.0, however, this
expansion can be disabled, and the tracking algorithms will walk the CSG trees verbatim, albeit at some tracking efficiency cost.  Therefore, most
output generated from the GDML to FLUKA conversion described here can only be used with FLUKA 4.0.

@@

\section{STL and CAD to GDML conversion}
STL and CAD conversion are closely related. In both cases the solid
(in the case of STL) and solids (in the case of CAD) are converted to
tessellated solid(s). STL is a relatively simple file format that can
be loaded using pure Python. As STL files typically only contain a single
solid the \pyinline{stl.Reader} provides a single solid and not a logical volume
as with the other file readers.
%
\begin{figure}
\begin{center}
\includegraphics[width=0.9\columnwidth]{SectorBend}
\caption{Example conversion of a simple CAD (STEP) geometry (sector bend
  dipole electromagnet) to GDML, above: the original model in FreeCAD,
  below: the GDML geometry viewed using \PYGEOMETRY{}. }
\label{fig:cad-gdml}
\end{center}
\end{figure}

@@

\section{Complete simulation example}
\PYGEOMETRY{} is designed to be as flexible as possible and offer the user a wide range of usage styles, input files
and workflows. A fictitious beamline was created to demonstrate the capabilities of \PYGEOMETRY, this creates a composite
\emph{scene} which consists of geometry sources from the different formats described in this paper. The beamline consists of a vacuum chamber
(modelled in \PYGEOMETRY{}), a vacuum gate valve (STL from the manufacturer), a triplet of quadrupole magnets (exported to GDML
from BDSIM), a sector bend dipole electromagnet (created in Autodesk Fusion 360) and finally a Faraday cup (FLUKA geometry designed in
flair). Each different file is loaded using the appropriate \PYGEOMETRY{} \pyinline{Reader} class and then placed as a physical volume. The final composite
geometry is shown in Figure~\ref{fig:model}. It must be noted that when this complete geometry is written to GDML and loaded into
Geant4 it cannot be visualised with anything but the ray tracer due to limitations in the OpenGL visualisation in Geant4. Another
important note is that this geometry cannot be converted to FLUKA as it contains tessellated solids (both the STL gate valve and
dipole magnet).
%
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{model.pdf}
\caption{Complete composite example using \PYGEOMETRY{} showing a model with geometry sourced from a range of different formats.}
\label{fig:model}
\end{center}
\end{figure*}

@@

\section{Author contributions}
{\bf Stewart Boogert}: conceptualization, supervision, methodology, software, validation and writing - original draft. 
{\bf Stuart Walker}: methodology, software, validation and writing - original draft (FLUKA components). 
{\bf Andrey Abramov}: software, validation and writing - review \& editing (Geant4 and GDML). 
{\bf Laurence Nevay}: software, validation validation and writing - review \& editing. 
{\bf William Shields}: validation and writing - review \& editing
